---
title: "PA1_template"
author: "Tarig Elamin"
date: "Friday, February 13, 2015"
output:
  md_document
    
---
---
                **Data Science Specialization**  Winter 2015

                **Reproduciable Research Course** 

                *****Assignment_1****************
                             
                by **Tarig Elamin*****
----

**in this assignment our plan is to answer  the following questions:**

1. What is mean total number of steps taken per day?

2. What is the average daily activity pattern?

3. Which 5-minute interval, on average across all the days in the dataset, 

contains the maximum number of steps?
 
4. Do these values differ from the estimates from the first part of the 

assignment?

5. What is the impact of imputing    

#####data access

The data downloaded from this link:

https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

and then unzipped in our working directory.Then, in the following steps below

,R program  code is built and run to find out about the answers.


                   **working directory** 
  C://Users//abdu//Desktop//data_science//reproducible research course

----

```{r}
activ <- read.csv("C://Users//abdu//Desktop//data_science//reproducible research course//activity.csv" ,colClasses = c("integer", "Date", "factor"))
activ$month <- as.numeric(format(activ$date, "%m"))
# ignoring the missing observations in this stage of analysis
noNA<- na.omit(activ)
rownames(noNA) <- 1:nrow(noNA )
head(noNA)
dim(noNA)
library(ggplot2)
```

** For the mean total number of steps taken per day and the median as well,**

**the R code is below**

```{r}
library(ggplot2)
library(gridExtra)
my_grob = grobTree(textGrob("Tarig Elamin,    project1!", x=0.1,  y=0.95, hjust=0,
  gp=gpar(col="yellow", fontsize=15, fontface="italic")))
z<-ggplot(noNA, aes(date, steps)) + geom_bar(stat = "identity", colour = "steelblue", fill = "springgreen", width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Graph of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")+annotation_custom(my_grob)
z + theme(panel.background = element_rect(fill = "blue"))
totalSteps <- aggregate(noNA$steps, list(Date = noNA$date), FUN = "sum")$x
mean(totalSteps)
median(totalSteps)
```

** resulting in :the mean total number of steps taken per day = **10766.19**

** resulting in :the median total number of steps taken per day = **10765**

----

**let's find out about the average daily activity pattern:**

```{r}

avgSteps <- aggregate(noNA$steps, list(interval = as.numeric(as.character(noNA$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"
library(gridExtra)
my_grob = grobTree(textGrob("Tarig Elamin,           project1!", x=0.1,  y=0.95, hjust=0,
  gp=gpar(col="blue", fontsize=15, fontface="italic")))
w<-ggplot(avgSteps, aes(interval, meanOfSteps)) + geom_line(color = "green3", size = 0.8) + labs(title = "Time Series graph of the 5_Minute Interval", x = "5_Minute Intervals", y = "Average_Number_ of _Steps Taken")+
  annotation_custom(my_grob)
w + theme(panel.background = element_rect(fill = "yellow"))

```

----

Across all the days in the activity dataset -on average,the 5_minute interval

, containing the maximum number of steps can be seen from this code below

```{r}
avgSteps[avgSteps$meanOfSteps == max(avgSteps$meanOfSteps), ]
```

so with the corresponding interval,the average of steps is turned out to be =**206.1698**

----

missing data represents a big problem if we attend to a regression model for 

example but first of all let's find out about the total number of missing 

observations/rows and decide which approach one can use to impute the missing 

data(one can select the mean or median or zero or a constant).Let's select 

the mean to impute our missing values. 




```{r}
sum(is.na(activ))
```

####the mean of 5_minute interval to replace the missing values in the new 

####data:

```{r}
newData <- activ 
for (i in 1:nrow(newData)) {
    if (is.na(newData$steps[i])) {
        newData$steps[i] <- avgSteps[which(newData$interval[i] == avgSteps$interval), ]$meanOfSteps
    }
}

head(newData)
sum(is.na(newData))
```

------

1. **graph of the total number of steps taken each day**
2. **Mean total number of steps taken per day**
3. **Median total number of steps taken per day**

 can be seen, below, for the new data that has no missing values:

```{r}
library(ggplot2)
ww<-ggplot(newData, aes(date, steps)) + geom_bar(stat = "identity",
                                             colour = "green3",
                                             fill = "green",
                                             width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Graph of Total Number of Steps Taken Each Day (NOmissing data)", x = "Date", y = "Total_number_of_steps")
ww+theme(panel.background = element_rect(fill = "blue"))

```


**Looking for mean and median of the old activ data(the one have missing** 

**values)against the new data that has nomissing.**

============================================================================

```{r}
newTotalSteps <- aggregate(newData$steps, 
                           list(Date = newData$date), 
                           FUN = "sum")$x
newMean <- mean(newTotalSteps)
newMean
newMedian <- median(newTotalSteps)
newMedian
oldMean <- mean(totalSteps)
oldMedian <- median(totalSteps)
newMean - oldMean
newMedian - oldMedian
```

 
Noting that there is no difference between the old comparing to the new mean

,but the new median is bigger than the old median.

---------

*The differences-if any- in activity patterns between weekdays and weekends*

```{r}
head(newData)
newData$weekdays <- factor(format(newData$date, "%A"))
levels(newData$weekdays)
levels(newData$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday", 
                                             "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(newData$weekdays)
table(newData$weekdays)
```


------

**The time series panel plot the 5_minute interval  and the average number**  

** of steps taken averaged across all weekday days or weekend days .**  

```{r}
avgSteps <- aggregate(newData$steps, 
                      list(interval = as.numeric(as.character(newData$interval)), 
                           weekdays = newData$weekdays),
                      FUN = "mean")
names(avgSteps)[3] <- "meanOfSteps"
library(lattice)
xyplot(avgSteps$meanOfSteps ~ avgSteps$interval |avgSteps$weekdays, 
        layout = c(1, 2), type = "l",
       xlab = "Interval", ylab = "Number of steps",main="Average all weekday or weekend ,\nby Tarig Elamin-project_1",pch=19,col="green3")

```

                                 **Appendix:*

**data definition:**

*variable name:definition*

1. steps: Number of steps taking in a 5-minute interval (missing values are 

coded 

as NA)

2. date: The date on which the measurement was taken in YYYY-MM-DD format

3. interval: Identifier for the 5-minute interval in which measurement was 

taken

   


                            ** The End **

